---
title: "Modules in R"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Modules in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, results='asis', echo=FALSE}
cat(gsub("\\n   ", "", packageDescription("modules", fields = "Description")))
```

## Installation

From CRAN:
```{r, eval=FALSE}
install.packages("modules")
```

From GitHub:

```{r, eval=FALSE}
devtools::install_github("wahani/modules")
```

# Introduction

The key idea of this package is to provide a unit which is self contained, i.e. 
has it's own scope. The main and most reliable infrastructure for such 
organizational units of source code in the R ecosystem is a package. Compared to
a package modules can be considered ad hoc, but - in the sense of an R-package -
self contained. Furthermore modules typically consist of one file; in contrast
to a package which can wrap an arbitrary number of files.

There are two use cases. First when you use modules to develop scripts, which is
subject of this section; And then inside of packages where modules act more like
objects, as in object-oriented-programming. Outside of packages modules know
only of the base environment, i.e. within a module the base environment is the
only *package* on the *search path*. Also they are always represented as a list
inside R. Thus they can be treated as bags of functions.

In the following examples you will see the function `module` to define modules.
Typically you do not have to call that function explicitly but instead call
`use` to load a module into your current session.

```{r}
library(modules)
m <- module({
  boringFunction <- function() cat("boring output")
})

m$boringFunction()
```

Since they are isolated from the `.GlobalEnv` the following object `hey` can not
be found:

```{r error=TRUE}
hey <- "hey"
m <- module({
  isolatedFunction <- function() hey
})
m$isolatedFunction()
```


## Imports

If you rely on exported objects of packages you can refer to them explicitly
using `::`:

```{r}
m <- module({
  functionWithDep <- function(x) stats::median(x)
})
m$functionWithDep(1:10)
```

Or you can use `import` for *attaching* single objects or packages and `use` for
*attaching* or loading a module:

```{r}
m <- module({
 
  import(stats, median) # make median from package stats available
  
  functionWithDep <- function(x) median(x)

})
m$functionWithDep(1:10)
```

```{r}
m <- module({
  
  import(stats)
  
  functionWithDep <- function(x) median(x)

})
m$functionWithDep(1:10)
```

## Exports

It may also be of interest to control which objects are visible for the client.
You can do that with the `export` function. Note that export accepts regular
expressions which are indicated by a leading '^'.

```{r}
m <- module({
  
  export("fun")
  
  .privateFunction <- identity
  privateFunction <- identity
  fun <- identity
  
})

names(m)
```

## Example: Modules as Parallel Process

One example where you may want to have more control of the enclosing environment 
of a function is when you parallelize your code. First consider the case when a 
*naive* implementation fails.

```{r error=TRUE}
library(parallel)
dependency <- identity
fun <- function(x) dependency(x) 

cl <- makeCluster(2)
clusterMap(cl, fun, 1:2)
stopCluster(cl)
```

To make the function `fun` self contained we can define it in a module. 

```{r}
m <- module({
  dependency <- identity
  fun <- function(x) dependency(x) 
})

cl <- makeCluster(2)
clusterMap(cl, m$fun, 1:2)
stopCluster(cl)
```


# Scripts as modules

You can make scripts into modules with `as.module` or implicitly when you refer 
to a file (or directory) in a call to `use`. Inside such a script you can use
`import` and `use` in the same way you typically use `library`. A major
difference is, that library will not only attach the stated package but also all
packages in the depends field of that dependency. This is something you have to
do manually (explicitly) with `import`. Consider the following example where we
create a module in a temporary file with its dependencies.

```{r}
code <- "
import(methods)
import(aoos)
# This is an example where we rely on functions in 'aoos':
list : generic(x) %g% standardGeneric('generic')
generic(x ~ ANY) %m% as.list(x)
"

fileName <- tempfile(fileext = ".R")
writeLines(code, fileName)
```

Then we can load such a module into this session by the following:

```{r}
someModule <- use(fileName)
someModule$generic(1:2)
```


# Documentation

If you want proper documentation for your functions or modules you really want a
package. However, there are some simple things you can do for ad-hoc
documentation of modules which is to use comments:

```{r}
m <- module({
  fun <- function(x) {
    ## A function for illustrating documentation
    ## x (numeric)
    x
  }
})
```

There are print methods for modules and functions within modules:

```{r}
m
m$fun
```


# Nested Modules

You can also write nested modules, which means you define modules inside
modules. In this case dependencies of the top level module are accessible to its
children:

```{r}
m <- module({
  
  import("stats", "median")
  import("modules", "module")
  
  anotherModule <- module({
    fun <- function(x) median(x)
  })
  
})

m$anotherModule$fun(1:2)
```


# Modules in Packages

When you use modules inside a R-package the search path of a module connects to 
the packages namespace. So it sees all objects within the package and has access
to all its dependencies. You can always change this by specifying the argument 
`topEncl`. Do not use `import` inside this setting but instead rely on the 
package to handle your dependencies. Attaching other modules using `use` and 
`expose` can make sense but is a matter of preference. You should definitely not
rely on modules in files and load them with `use`!


# Modules with Object Orientation

## S3

S3 method dispatch will not work because of the special search mechanism of 
`UseMethod`. What will work, however, is attaching the module so `UseMethod` can
find the methods.

```{r}
m <- module({
  generic <- function(x) UseMethod("generic")
  generic.numeric <- function(x) cat("method for x ~ numeric")
})
# m$generic(1) # this won't work
use(m, attach = TRUE)
m$generic(1)
```

## S4

More reliable is the dispatch in S4. By default the set functions of the methods
package have side effects in the top level environment. So you would have to set
the appropriate environment for the argument 'where'. 'aoos' provides syntactic 
sugar (S4) and has side effects in the environment in which the constructor 
functions are called so method dispatch will work without extra work:

```{r}
m <- module({
  import("methods")
  import("aoos")
  gen(x) %g% cat("default method")
  gen(x ~ numeric) %m% cat("method for x ~ numeric")
})
m$gen("Hej")
m$gen(1)
```

S4 classes (or types in aoos) have side effects in the global environment and S4
classes are searched for in the top level environment. This means that the
module is not self-contained because the class definition will be cached
somewhere else.

```{r}
m <- module({
  import("methods")
  import("aoos")
  numeric : NewType() %type% .Object
})
m$NewType(1)
```


# Modules and Coupling

Best is to only use `::` and `use` with `attach = FALSE` to be explicit with 
your dependencies. However, there are some other options you have, which will
result in stronger forms of coupling between modules.

## Modules to Model Mutable State

This is in itself abstract but in principle you can not only put functions in
your bag (module) but any R-object. This transforms into something which is
probably associated with object-orientation. You can like this or not, here I
simply use it to illustrate the strongest form of coupling between two modules I
can come up with. 

In the following I define a module to encapsulate some value and have a *get*
and *set* method for it:

```{r}
mutableModule <- module({
  .num <- NULL
  get <- function() .num
  set <- function(val) .num <<- val
})
mutableModule$get()
mutableModule$set(2)
```

## Coupling Between Modules

In the next module we can use `mutableModule` and rebuild the interface to
`.num`.

```{r}
complectModule <- module({
  use(.GlobalEnv$mutableModule, attach = TRUE)
  getNum <- function() get()
  set(3)
})
mutableModule$get()
complectModule$getNum()
```

Depending on your expectations with respect to the above code it comes at a
surprise that we can get and set that value from an attached module; Furthermore
it is not changed in `mutableModule`. This is because `use` will trigger a
re-initialization of any module you plug in. You can override this behaviour:

```{r}
complectModule <- module({
  use(.GlobalEnv$mutableModule, attach = TRUE, reInit = FALSE)
  getNum <- function() get()
  set(3)
})
mutableModule$get()
complectModule$getNum()
```

This is not all we can do. Also we can use `expose`. This function will take
everything in a module and expose it to the environment from which it is called.

```{r}
complectModule <- module({
  expose(.GlobalEnv$mutableModule, reInit = TRUE)
  set(4)
})
mutableModule$get()
complectModule$get()
```

And of course we can do this with `reInit = FALSE` should this be desirable. In
this case both modules are essentially the same.

```{r}
complectModule <- module({
  expose(.GlobalEnv$mutableModule, reInit = FALSE)
  set(1)
})
mutableModule$get()
complectModule$get()
```


# Related Projects

There exist several projects with similar goals. First of all, the package
[klmr/modules](https://github.com/klmr/modules) aims at providing a unit similar
to what [Python](https://www.python.org/)-modules are. This projects is
obviously interesting for you when you have prior knowledge in Python. In
contrast to `klmr/modules` modules defined in this package do not aim for a full
replacement of R-packages -- rather they provide a sub-unit within R's package
ecosystem. Otherwise there is considerable overlap of features between the two
packages.

Second you may be interested in
[import](https://cran.r-project.org/package=import) which provides convenient
syntax for stating dependencies in script files. This is something which is also
covered here, although, when you are only interested in a replacement for
`library` the package `import` is more focused. 

`modules` in this package can act as objects as in object-orientation. In 
contrast to [R6](https://cran.r-project.org/package=R6) and reference classes
implemented in the methods package here these objects are immutable by default.
Furthermore it is not being made easy to change state of a module; but it is not
difficult to do that if you really want to: see the section on coupling.
Furthermore inheritance is not a feature, instead you have various possibilities
for object composition. If substitutability is really an issue then you may
also look at `aoos::retList`.

The development of the `modules` package has been inspired by other languages:
[F#](https://fsharpforfunandprofit.com/posts/organizing-functions/), 
[Erlang](http://learnyousomeerlang.com/modules) and 
[julia](http://docs.julialang.org/en/release-0.4/manual/modules/).

